---
title: "Reproducible Research: Peer Assessment 1"
author: "HvdR"
date: "Friday, July 17, 2015"
output: 
  html_document:
    keep_md: true
---
```{r setting,echo=FALSE}
setwd("H:/Coursera-Datascientist/05-RepData/Assignment1/git-repo")
```
## a) Introduction

This document describes the analysis of the data about personal 
movement during the months of October and November, 2012.

## b) Loading and preprocessing the data
The dataset is copied from 
https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip to "activity.zip". The data is read in with:
```{r b-reading}
activity1 <- read.csv(unz("activity.zip", "activity.csv"))
```
The data consists of: 

- date: the date of the measurement
- interval: the 5-minute interval of the day
- steps: the number of steps taken in the 5-minute interval

## c) What is mean total number of steps taken per day?
The total number of steps taken per day is calculated with:
```{r c-total}
stepsperday1 <- aggregate(steps ~ date, data = activity1, sum)
```
The histogram of the total steps taken per day is:
```{r c-histogram,fig.height=4}
hist(stepsperday1$steps, 
     col="lightblue", 
     main="Activity data",
     xlab="Steps per day",
     ylab="Frequency")
```

Calculating the mean and median of the total number of steps taken per day is done with the summary function:

```{r c-summary}
summarystepsperday1 <- summary(stepsperday1$steps)
daymedian1 <- as.integer(summarystepsperday1["Median"])
daymean1 <- as.integer(summarystepsperday1["Mean"])
daytotal1 <- as.integer(sum(stepsperday1$steps, na.rm=TRUE))
```

The mean and median of the total number of steps taken per 
day is:

- Mean: **`r daymean1`**.
- Median: **`r daymedian1`**.


## d) What is the average daily activity pattern?

The average daily activity pattern consists of the mean of 
each 5-minute interval accross all days in the dataset.

The calculation is done with:

```{r d-interval}
stepsperinterval1 <- aggregate(steps ~ interval, data = activity1, mean )
```

The plotting is: 
```{r d-intervalplot,fig.height=4}
plot(stepsperinterval1$interval, 
     stepsperinterval1$steps, 
     xlab= "Day-interval (5-minute)", 
     ylab= "Average nr of steps taken", 
     type='l', 
     col='red')
title(main="Averaged across all days", 
      col.main="black", 
      font.main=4)
```

The 5-minute interval accross all days with the maximum mean value is caculated with:

```{r d-intervalmax}
intervalmax1 <- (stepsperinterval1[ 
   stepsperinterval1$steps == max( stepsperinterval1$steps),1 ])
```

The 5-minute interval accross all days with the maximum mean value is: 

- Interval with maximum mean: **`r intervalmax1` **.

## e) Imputing missing values

Calculating the total number of missing values in the dataset:

```{r e-missing}
missingvalues <- sum(is.na(activity1$steps))
totalvalues <- nrow(activity1)
```

The missing and total number of values is:

- Total number of values: **`r totalvalues`**.
- Number of missing values: **`r missingvalues`**.

The strategy for filling in all of the missing values is:
*calculate the rounded mean of the 5-minute interval of the other days*. 

```{r e-replacemissing}
replacena <- function(x,interval) {
  if (is.na(x))
    round(stepsperinterval1[
      stepsperinterval1$interval == interval,][,2])
  else
    x
}
```

Creating a new dataset with missing data filled in:
```{r e-makedataset}
# copy dataset
activity2 <- activity1
# repalce na's
activity2$steps <- mapply(replacena,
                          activity2$steps,activity2$interval)
```

Showing the histogram with missing data filled in:
```{r e-comparehistogram,fig.height=4}

# the histogram of the date with missing data filled
hist(stepsperday2$steps, 
     col="lightgreen", 
     main="Corrected activity data",
     xlab="Steps per day",
     ylab="Frequency")

```

Calculation of the new median and mean values:
```{r e-summary}
summarystepsperday2 <- summary(stepsperday2$steps)
daymedian2 <- as.integer(summarystepsperday2["Median"])
daymean2 <- as.integer(summarystepsperday2["Mean"])
daytotal2 <- as.integer(sum(stepsperday2$steps))
```

The mean and median of the total number of steps taken per 
day with the missing values filled in is:

- Mean: **`r daymean2`**.
- Median: **`r daymedian2`**.

### Consequences of filling in the missing values

- The "Mean" and "Median" values don't change: 
using the mean values of the other days doesn't 
disturb the distribution

- The "Total" values of the do change from 
original **`r daytotal1`** to **`r daytotal2`**:
replacing missing values with mean values results
in more steps.


## Are there differences in activity patterns between weekdays and weekends?
